Periodograms for Multiband Astronomical Time Series

### 2015-02-05 writeup/paper.tex
Change to 2 column w/ emulateapj

\documentclass[12pt,preprint]{aastex}

@@

Because astronomical observing cadences are rarely so uniform, many have looked at extending the ideas behind the periodogram to work with unevenly-sampled data. Most famously, \citet{Lomb76} and \citet{Scargle82} extended earlier work to define the {\it normalized periodogram}:
\begin{equation}
  \eqlabel{LombScargle}
  P_N(\omega) = \frac{1}{2\,V_y}\left[
    \frac{\left[\sum_k(y_k - \bar{y})\cos\omega(t_k - \tau)\right]^2}
    {\sum_k \cos^2\omega(t_k - \tau)}
    +
    \frac{\left[\sum_k(y_k - \bar{y})\sin\omega(t_k - \tau)\right]^2}
    {\sum_k \sin^2\omega(t_k - \tau)}
\right],
\end{equation}
where $\bar{y}$ is the mean and $V_y$ is the variance of the data $\{y_k\}$, and $\tau$ is the time-offset which makes $P_N(\omega)$ independent of a translation in $t$ \citep[see][for an in-depth discussion]{NumRec}. \citet{Lomb76} showed that this time-offset has a deeper effect: namely, it makes $P_N$ identical to the estimate of harmonic content given a least-squares fit to a single-component sinusoidal model,
\begin{equation}
  \eqlabel{SingleModel}
  d(t) = A\sin(\omega t + \phi).
\end{equation}
This long-recognized connection between spectral power and least squares fitting methods was solidified by \citet{Jaynes87}, who demonstrated that the normalized periodogram of Lomb and Scargle is a sufficient statistic for inferences about a stationary-frequency signal in the presence of Gaussian noise. Building on this result, \citet{Bretthorst88} explored the extension of these methods to more complicated models with multiple frequency terms, non-stationary frequencies, and other more sophisticated models within a Bayesian framework.

@@

\subsection{Matrix Formalism}
The expressions related to the stationary sinusoid model can be expressed more compactly by defining the following matrices:
\begin{equation}
X_\omega = \left[
\begin{array}{cc}
\sin(\omega t_1) & \cos(\omega t_1)\\
\sin(\omega t_2) & \cos(\omega t_2)\\
\vdots & \vdots \\
\sin(\omega t_N) & \cos(\omega t_N)\\
\end{array}
\right];~~
y = \left[
\begin{array}{c}
y_1 \\
y_2\\
\vdots \\
y_N\\
\end{array}
\right];~~
\Sigma = \left[
\begin{array}{cccc}
\sigma_1^2 & 0 &  \cdots & 0\\
0 & \sigma_2^2 &  \cdots & 0\\
\vdots & \vdots &  \ddots & \vdots\\
0 & 0 &  \cdots & \sigma_N^2
\end{array}
\right]
\end{equation}
With these definitions, the model in \eq{simplemodel} can be expressed as a simple linear product, $y(t|\omega,\theta) = X_\omega\theta$, and the model and reference $\chi^2$ can be written

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig01.pdf}
  \caption{
    An illustration of the basic periodogram and its relationship to the single-term sinusoid model. The left panel shows the input data, while the right panels show the fit derived from the data. The upper-right panel shows the periodogram with a clear peak at the true period of 0.622 days, and the bottom-right panel shows the data as a function of the phase associated with this period. Note in the periodogram the presence of the typical aliasing effect, with power located at beat frequencies between the true period and the 1-day observing cadence (see \sect{simple_period} for further discussion).
  }
  \figlabel{basic_example}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig02.pdf}
  \caption{
    An illustration of the effect of the floating mean model for censored data.
    The data consist of 80 observations drawn from a sinusoidal model. To mimic a potentially damaging selection effect, all observations with magnitude fainter than 16 are removed (indicated by the light-gray points). The standard and floating-mean periodograms are computed from the remaining data; these fits are shown over the data in the left panel. Because of this biased observing pattern, the mean of the observed data is a biased estimator of the true mean. The standard fixed-mean model in this case fails to recover the true period of 0.622 days, while the floating mean model still finds the correct period.
  }
  \figlabel{floating_mean}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig03.pdf}
  \caption{
    The model fits and periodograms for several truncated Fourier models.
    The data are the same as those in \fig{basic_example}. Note that the
    higher-order models will generally show periodogram peaks at multiples
    of the true fundamental frequency $P_0$: this is because for integer $n$
    less than the number of Fourier terms in the model, $P_0$ is a higher
    harmonic of the model at $P=nP_0$. Additionally, the increased degrees of
    freedom in the higher-order models let them fit better at any frequency,
    which drives up the ``background'' level in the periodogram.
  }
  \figlabel{multiterm_example}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig04.pdf}
  \caption{
    The effect of regularization on a high-order model. The data is the same as
    those in \fig{basic_example}. We fit a 20-term truncated Fourier model to
    the data, with and without a regularization term. Without regularization,
    the model oscillates widely to fit the noise in the data. The
    regularization term effectively damps the higher-order Fourier modes and
    removes this oscillating behavior, leading to a more robust model with
    stronger periodogram peaks.
  }
  \figlabel{regularized_example}
\end{figure}

@@

\section{A Multiple-Band Model}
\sectlabel{multiband}
In this section we will apply what was learned in the previous sections to construct the {\it multiband periodogram} which can flexibly account for heterogeneous sources of data for a single object.
To compute such a periodogram, we will take advantage of the easy extensibility of the matrix formalism which led to our generalizations above.
The multiband model contains the following features:
\begin{enumerate}
  \item An $N_{base}$-term truncated Fourier fit which models a latent parameter, which we'll call the ``overall variability''.
  \item A set of $N_{band}$-term truncated Fourier fits, each of which models the residual of a single band from this overall variability.
\end{enumerate}
The total number of parameters for $K$ filters is then $M_K = (2N_{base} + 1) + K(2N_{band} + 1)$. As a result, for each band $k$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  \eqlabel{multiband_model}
  y_k(t|\omega,\theta) = &\theta_0 + \sum_{n=1}^{M_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right]& +\nonumber\\ 
  &\theta^{(k)}_0 + \sum_{n=1}^{M_{band}} \left[\theta^{(k)}_{2n - 1}\sin(n\omega t) + \theta^{(k)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(k)}$ are determined individually.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig05a.pdf}
  \includegraphics[width=\textwidth]{fig05b.pdf}
  \caption{
    An illustration of the performance of the multiband periodogram. The
    upper panels show simulated $ugriz$ observations of an RR Lyrae light
    curve in which all 5 bands are observed each night. With 60 observations
    in each band, a periodogram computed from any single band is sufficient to
    determine the true period of 0.622 days. The lower panels show the same
    data, except with only a single $ugriz$ band observed each night (i.e.
    12 observations per band). In this case, no single band has enough
    information to detect the period. The shared-phase multiband approach
    of \sect{multiband} (lower-right panel) combines the information from
    all five bands, and results in a significant detection of the true period.
  }
  \figlabel{multiband_sim}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{fig06.pdf}
  \caption{
    Comparison of the periodograms produced by various multiband models.
    The data is the same as that used in \fig{multiband_sim}. $N_{base}$ gives
    the number of Fourier terms in the base model, and $N_{band}$ gives the
    number of Fourier terms used to fit the residuals around this model within
    each band. The characteristics discussed with previous figures are also
    seen here: in particular, the level of ``background noise'' in the
    periodogram grows with the model complexity $M$,
  } 
  \figlabel{multiband_models}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig07a.pdf}
  \includegraphics[width=\textwidth]{fig07b.pdf}
  \caption{
    Comparison of the Multiband algorithm and single-band supersmoother algorithm on 483 well-sampled RR Lyrae light curves from Stripe 82.
    The upper panels show a representative lightcurve and periodogram fits, while the bottom panels compare the derived periods to the template-based periods reported in S10.
    Shown for reference are the beat aliases (dotted lines) and the multiplicative alias (dashed lines): numbers along the top and right edges of the panels indicate the number of points aligned with each trend.
    The single-band supersmoother model tends to err toward multiplicative aliases, while the multiband model tends to err toward beat frequency aliases.
    Both methods find the correct period among the top 5 significant peaks around 99\% of the time.
  } 
  \figlabel{compare_periods}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig08a.pdf}
  \includegraphics[width=\textwidth]{fig08b.pdf}
  \caption{
    This figure repeats the experiment shown in \fig{compare_periods} (see caption there for description), but the data is artificially reduced to only a single-band observation on each evening, a situation reflective of the observing cadence of future large-scale surveys.
    In this case, the single-band SuperSmoother strategy used as a first pass in S10 fails: there is simply not enough data in each band to recover an accurate period estimate. The correct period is among the top 5 candidates in fewer than 50\% of cases.
    The shared-phase multiband approach utilizes information from all five bands, and returns much more robust results: even with the greatly-reduced data, the true period is among the top 5 candidates in 93\% of cases.
  } 
  \figlabel{compare_periods_reduced}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{fig09.pdf}
  \caption{
    Fraction of periods correctly determined for LSST RR Lyrae as a function
    of the length of the observing season and the mean $g$-band magnitude, for the multiband (solid lines)
    and single-band supersmoother (dashed lines) approaches.
    The multiband method is superior to the single-band supersmoother approach in all cases, and especially for the faintest objects.
  } 
  \figlabel{LSST_sims}
\end{figure}

@@

\appendix
\section{Python Implementation of Multiband Periodogram}
\sectlabel{gatspy}
The algorithm outlined in this paper is available in {\tt gatspy}, an open-source Python package for general astronomical time-series analysis\footnote{\url{http://github.com/astroml/gatspy/}} \citep{gatspy}. Along with the periodogram implementation, it also contains code to download all the data used in this work. Code to reproduce this paper, including all figures, is available in a separate repository\footnote{\url{http://github.com/jakevdp/multiband_LS/}}.

@@

\section{Comparison with \citet{Long2014}}
\sectlabel{long_comparison}
As we were finishing this study, we learned that another group had released a preprint independently addressing the multiband periodogram case, and come up with a solution very similar to the one presented here \citep[][hereafter LCB14]{Long2014}.
They present two methods, the ``Multiband Generalized Lomb-Scargle'' (MGLS) which is effectively identical to the $(1, 0)$ multi-phase model
here, and the ``Penalized Generalized Lomb-Scargle'' (PGLS), which is similar in spirit to our $(0, 1)$ shared-phase model.

>>>>

\documentclass{emulateapj}

@@

Because astronomical observing cadences are rarely so uniform, many have looked at extending the ideas behind the periodogram to work with unevenly-sampled data. Most famously, \citet{Lomb76} and \citet{Scargle82} extended earlier work to define the {\it normalized periodogram}:
\begin{eqnarray}
  \eqlabel{LombScargle}
  P_N(\omega) = \frac{1}{2\,V_y}
  \Bigg[&\frac{\left[\sum_k(y_k - \bar{y})\cos\omega(t_k - \tau)\right]^2}
    {\sum_k \cos^2\omega(t_k - \tau)} +\nonumber\\
   & \frac{\left[\sum_k(y_k - \bar{y})\sin\omega(t_k - \tau)\right]^2}
    {\sum_k \sin^2\omega(t_k - \tau)}\Bigg],
\end{eqnarray}
where $\bar{y}$ is the mean and $V_y$ is the variance of the data $\{y_k\}$, and $\tau$ is the time-offset which makes $P_N(\omega)$ independent of a translation in $t$ \citep[see][for an in-depth discussion]{NumRec}. \citet{Lomb76} showed that this time-offset has a deeper effect: namely, it makes $P_N$ identical to the estimate of harmonic content given a least-squares fit to a single-component sinusoidal model,
\begin{equation}
  \eqlabel{SingleModel}
  d(t) = A\sin(\omega t + \phi).
\end{equation}
This long-recognized connection between spectral power and least squares fitting methods was solidified by \citet{Jaynes87}, who demonstrated that the normalized periodogram of Lomb and Scargle is a sufficient statistic for inferences about a stationary-frequency signal in the presence of Gaussian noise. Building on this result, \citet{Bretthorst88} explored the extension of these methods to more complicated models with multiple frequency terms, non-stationary frequencies, and other more sophisticated models within a Bayesian framework.

@@

\subsection{Matrix Formalism}
The expressions related to the stationary sinusoid model can be expressed more compactly by defining the following matrices:
\begin{eqnarray}
X_\omega = \left[
\begin{array}{cc}
\sin(\omega t_1) & \cos(\omega t_1)\\
\sin(\omega t_2) & \cos(\omega t_2)\\
\vdots & \vdots \\
\sin(\omega t_N) & \cos(\omega t_N)\\
\end{array}
\right]; \nonumber\\
y = \left[
\begin{array}{c}
y_1 \\
y_2\\
\vdots \\
y_N\\
\end{array}
\right];~~
\Sigma = \left[
\begin{array}{cccc}
\sigma_1^2 & 0 &  \cdots & 0\\
0 & \sigma_2^2 &  \cdots & 0\\
\vdots & \vdots &  \ddots & \vdots\\
0 & 0 &  \cdots & \sigma_N^2
\end{array}
\right]
\end{eqnarray}
With these definitions, the model in \eq{simplemodel} can be expressed as a simple linear product, $y(t|\omega,\theta) = X_\omega\theta$, and the model and reference $\chi^2$ can be written

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig01.pdf}
  \caption{
    An illustration of the basic periodogram and its relationship to the single-term sinusoid model. The left panel shows the input data, while the right panels show the fit derived from the data. The upper-right panel shows the periodogram with a clear peak at the true period of 0.622 days, and the bottom-right panel shows the data as a function of the phase associated with this period. Note in the periodogram the presence of the typical aliasing effect, with power located at beat frequencies between the true period and the 1-day observing cadence (see \sect{simple_period} for further discussion).
  }
  \figlabel{basic_example}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig02.pdf}
  \caption{
    An illustration of the effect of the floating mean model for censored data.
    The data consist of 80 observations drawn from a sinusoidal model. To mimic a potentially damaging selection effect, all observations with magnitude fainter than 16 are removed (indicated by the light-gray points). The standard and floating-mean periodograms are computed from the remaining data; these fits are shown over the data in the left panel. Because of this biased observing pattern, the mean of the observed data is a biased estimator of the true mean. The standard fixed-mean model in this case fails to recover the true period of 0.622 days, while the floating mean model still finds the correct period.
  }
  \figlabel{floating_mean}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig03.pdf}
  \caption{
    The model fits and periodograms for several truncated Fourier models.
    The data are the same as those in \fig{basic_example}. Note that the
    higher-order models will generally show periodogram peaks at multiples
    of the true fundamental frequency $P_0$: this is because for integer $n$
    less than the number of Fourier terms in the model, $P_0$ is a higher
    harmonic of the model at $P=nP_0$. Additionally, the increased degrees of
    freedom in the higher-order models let them fit better at any frequency,
    which drives up the ``background'' level in the periodogram.
  }
  \figlabel{multiterm_example}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig04.pdf}
  \caption{
    The effect of regularization on a high-order model. The data is the same as
    those in \fig{basic_example}. We fit a 20-term truncated Fourier model to
    the data, with and without a regularization term. Without regularization,
    the model oscillates widely to fit the noise in the data. The
    regularization term effectively damps the higher-order Fourier modes and
    removes this oscillating behavior, leading to a more robust model with
    stronger periodogram peaks.
  }
  \figlabel{regularized_example}
\end{figure*}

@@

\section{A Multiple-Band Model}
\sectlabel{multiband}
In this section we will apply what was learned in the previous sections to construct the {\it multiband periodogram} which can flexibly account for heterogeneous sources of data for a single object.
To compute such a periodogram, we will take advantage of the easy extensibility of the matrix formalism which led to our generalizations above.
The multiband model contains the following features:
\begin{enumerate}
  \item An $N_{base}$-term truncated Fourier fit which models a latent parameter, which we'll call the ``overall variability''.
  \item A set of $N_{band}$-term truncated Fourier fits, each of which models the residual of a single band from this overall variability.
\end{enumerate}
The total number of parameters for $K$ filters is then $M_K = (2N_{base} + 1) + K(2N_{band} + 1)$. As a result, for each band $k$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  \eqlabel{multiband_model}
  &y_k(t|\omega,\theta) =
  \theta_0 + \sum_{n=1}^{M_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right] +&\nonumber\\ 
  &\theta^{(k)}_0 + \sum_{n=1}^{M_{band}} \left[\theta^{(k)}_{2n - 1}\sin(n\omega t) + \theta^{(k)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(k)}$ are determined individually.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig05a.pdf}
  \includegraphics[width=\textwidth]{fig05b.pdf}
  \caption{
    An illustration of the performance of the multiband periodogram. The
    upper panels show simulated $ugriz$ observations of an RR Lyrae light
    curve in which all 5 bands are observed each night. With 60 observations
    in each band, a periodogram computed from any single band is sufficient to
    determine the true period of 0.622 days. The lower panels show the same
    data, except with only a single $ugriz$ band observed each night (i.e.
    12 observations per band). In this case, no single band has enough
    information to detect the period. The shared-phase multiband approach
    of \sect{multiband} (lower-right panel) combines the information from
    all five bands, and results in a significant detection of the true period.
  }
  \figlabel{multiband_sim}
\end{figure*}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{fig06.pdf}
  \caption{
    Comparison of the periodograms produced by various multiband models.
    The data is the same as that used in \fig{multiband_sim}. $N_{base}$ gives
    the number of Fourier terms in the base model, and $N_{band}$ gives the
    number of Fourier terms used to fit the residuals around this model within
    each band. The characteristics discussed with previous figures are also
    seen here: in particular, the level of ``background noise'' in the
    periodogram grows with the model complexity $M$,
  } 
  \figlabel{multiband_models}
\end{figure}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig07a.pdf}
  \includegraphics[width=\textwidth]{fig07b.pdf}
  \caption{
    Comparison of the Multiband algorithm and single-band supersmoother algorithm on 483 well-sampled RR Lyrae light curves from Stripe 82.
    The upper panels show a representative lightcurve and periodogram fits, while the bottom panels compare the derived periods to the template-based periods reported in S10.
    Shown for reference are the beat aliases (dotted lines) and the multiplicative alias (dashed lines): numbers along the top and right edges of the panels indicate the number of points aligned with each trend.
    The single-band supersmoother model tends to err toward multiplicative aliases, while the multiband model tends to err toward beat frequency aliases.
    Both methods find the correct period among the top 5 significant peaks around 99\% of the time.
  } 
  \figlabel{compare_periods}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig08a.pdf}
  \includegraphics[width=\textwidth]{fig08b.pdf}
  \caption{
    This figure repeats the experiment shown in \fig{compare_periods} (see caption there for description), but the data is artificially reduced to only a single-band observation on each evening, a situation reflective of the observing cadence of future large-scale surveys.
    In this case, the single-band SuperSmoother strategy used as a first pass in S10 fails: there is simply not enough data in each band to recover an accurate period estimate. The correct period is among the top 5 candidates in fewer than 50\% of cases.
    The shared-phase multiband approach utilizes information from all five bands, and returns much more robust results: even with the greatly-reduced data, the true period is among the top 5 candidates in 93\% of cases.
  } 
  \figlabel{compare_periods_reduced}
\end{figure*}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{fig09.pdf}
  \caption{
    Fraction of periods correctly determined for LSST RR Lyrae as a function
    of the length of the observing season and the mean $g$-band magnitude, for the multiband (solid lines)
    and single-band supersmoother (dashed lines) approaches.
    The multiband method is superior to the single-band supersmoother approach in all cases, and especially for the faintest objects.
  } 
  \figlabel{LSST_sims}
\end{figure}

@@

\appendix
\section{Python Implementation of Multiband Periodogram}
\sectlabel{gatspy}
The algorithm outlined in this paper is available in {\tt gatspy}, an open-source Python package for general astronomical time-series analysis\footnote{\url{http://github.com/astroml/gatspy/}} \citep{gatspy}. Along with the periodogram implementation, it also contains code to download all the data used in this work. Code to reproduce this paper, including all figures, is available in a separate repository\footnote{\url{http://github.com/jakevdp/multiband\_LS/}}.

@@

\section{Comparison with Long (2014)}
\sectlabel{long_comparison}
As we were finishing this study, we learned that another group had released a preprint independently addressing the multiband periodogram case, and come up with a solution very similar to the one presented here \citep[][hereafter LCB14]{Long2014}.
They present two methods, the ``Multiband Generalized Lomb-Scargle'' (MGLS) which is effectively identical to the $(1, 0)$ multi-phase model
here, and the ``Penalized Generalized Lomb-Scargle'' (PGLS), which is similar in spirit to our $(0, 1)$ shared-phase model.
